The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-07-31 paper/referee/response-revision1.txt

MN-14-1011-MJ: Authors' response to the reviewer's comments
===========================================================

We are grateful to the referee for the insightful feedback, which has helped us
to improve the presentation of the data products.  In his/her report, the 
referee identified areas where additional information would (i) strengthen the 
role of the paper as a user's guide to the survey, and (ii) would lessen the 
burden on the reader to track down information from previous works.  
We appreciate the merits of the referee's suggestions in this respect, and have 
acted upon them by adding a total of 4 new pages (3 new sections, 3 new figures,
and 1 new table) to the manuscript.  Changes have been marked in bold.
We have also ~doubled the auxillary documentation that is available on our web
site a direct consequence of the referee's suggestions, and will continue to
expand the on line documentation as user experience accumulates.

@@

In what follows we will start by explaining our response to each of the issues 
which the referee identified as high-level remarks (9 items), followed by the
referee's comments on specific sections (8 items).

High-level items
----------------

[Referee]
> 1. This manuscript presents a large amount of information about IPHAS and the preparation
> of and contents of the data products.  It also relies heavily on previous
> publications for general descriptions of the survey, particulars of the instrumentation,
> data processing algorithms and earlier data releases.   As such, it puts considerable burden 
> on the readers to track down information from many different sources.  It is
> understandable that the authors prefer to not duplicate information that is already
> available elsewhere, but it will make it much easier to understand and use the IPHAS data
> products if at least some of the general information  could be repeated in this
> manuscript.
> For example, it would be helpful to include concise summary of the IPHAS survey
> parameters and this data release in tabular form.  This summary would serve as
> a "IPHAS-At-A-Glance" for users, and coule include the survey area, dates, telescope,
> camera and survey bandpasses, etc.  Much of the information is presented in the paper,
> but widely spread over the manuscript.  Without at least a guide to finding the
> information, users will be challenged to find what they need.

[Authors' response]
Whilst revising the manuscript in response to the items that follow below, we have 
kept the referee's recommendation to lessen the burden on the reader in mind.
As an immediate response to this specific item, we have extended Section 2 to 
(i) include an overview table of the survey parameters, summarising 15 key properties 
of the survey (Table 1), and (ii) included a plot reminding the reader of the footprint
of a single IPHAS pointing and its offset position (Figure 1).

@@

[Referee]
> 2. It was not immediately obvious if the images and catalogue from the Second IPHAS
> release supersede those in the 2008 Initial Data Release.  I believe the
> Second Release includes a reprocessing of all of the survey data to-date, so it
> is intended to replace the Initial Release, but that should be stated explicitly
> in the Introduction (e.g. in the 3rd paragraph).
> If my assumption is correct, there should be a summary statement about why the
> Second Release images and catalogue are in fact better products.  For example,
> better calibrations, processing algorithms, etc.

[Authors' response]
The referee correctly identified that we forgot to state that this new release 
supersedes the previous one.  We have added an explicit statement in this sense
to the third paragraph of Section 1.  We have also Section 3.1 to highlight 
the key differences between the past and the present release (last paragraph).

[Referee]
> 3. As stated in Section 7.2, the aim of IPHAS is to enable the discovery of
> emission line objects in the Galactic Plane.  Consequently, I was very surprised
> that no discussion of the sensitivity to line emission (e.g. H-alpha EW) was
> presented in the manuscript.  Perhaps this will be addressed further in the planned
> Barentsen et al paper, but it seems that at least some rudimentary demonstration
> of the sensitivity to line emission is a requirement for this paper.  The
> general photometric sensitivity in the three bands is discussed in section 6.2, and
> an illustration is given how to find emission-line candidates using the color-color
> diagram in Figure 18.  I strongly recommend that at least some examples be given
> of the r-H-alpha color excess as a function of line EW for a sample of objects
> with published spectra.

[Authors' response]
The referee requested a discussion on the sensitivity of the H-alpha 
photometry to line emission to be included in this work.  In response, we have
added one new section (7.4) and one new figure (Fig 21) to the paper.
In this new section, which is titled "r-Halpha excess as a quantitative measure
of Halpha emission", we compare the spectroscopic H-alpha EWs for a sample of
classical Be stars from Raddi et al (2014) against the position of these
objects in the IPHAS colour-colour diagram.  A good agreement is found.
Moreover, the accompanying discussion offers a guide to the previous works on 
this topic.

@@

[Authors' response]
We have edited Section 3.1 to offer the reader more details on the astrometric 
calibration.  This section now also includes the results of an external validation of 
our astrometry against UCAC4, which showed a mean residual of 0.09 +/- 0.07 arcsec at the 
bright end of our catalogue.  We have also added a statement to specify that
no proper motion information was used in arriving at our astrometry, which is
of limited effect because hundreds 2MASS positional reference stars were available 
in each CCD to fit the parameters of the WCS solution.

[Referee]
(There does not appear to be an item number 8 in the referee report.)
> 9. The authors point out in the introduction that IPHAS will complement a number of
> new infrared surveys of the Galactic Plane.  The paper would benefit from at least
> one cross-comparison with the catalogue of one of those surveys, such as GLIMPSE
> or MIPSGAL, as an illustration of how the combination of surveys will enable
> new science.  At present, the only tie-in to the infrared surveys is the
> visual comparison between IPHAS and GLIMPSE images of one field.

[Authors' response]
In response to the referee's suggestion to include a cross-comparison between
IPHAS and an infrared survey, we have added a new Section 7.3 which provides
a brief discussion on the overlap between the GLIMPSE and IPHAS catalogues.  
The discussion reminds the user that an optical survey will naturally be 
extinction-limited, by showing the sources which IPHAS and Spitzer do not have
in common as a function of their position on the Spitzer colour-colour diagram.

### 2014-07-31 paper/iphas-dr2-rev1.tex
First draft for resubmission

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

%This approach works very well 
%across the vast majority of the survey area.  Nevertheless,
%the Galactic Plane contains crowded regions 
%with large numbers of overlapping sources
%or rapidly spatially-varying nebulosity,
%where aperture photometry will be compromised by frequent blending
%or poor background subtraction.

Having carried out object detection and characterisation,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
which uses a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the field distortion of the WFC.
The parameters of this solution
are then progressively refined using the
Two-Micron All Sky Survey catalogue \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector
\citep{Gonzalez-Solares2008}.

An external validation has been carried out
by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the astrometric residuals
between IPHAS and UCAC4
for N million stars in the magnitude
range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean residual to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of position residuals
            between IPHAS DR2 and UCAC4.
            The residuals were computed for stars
            in the magnitude range $13 < r < 15$ which were
            not blended and not saturated.}
    \label{fig:astrometry}
\end{figure}

@@

In previous work, we have shown that the majority of
\ha-emitters seen by IPHAS towards an H{\sc ii} region
are likely to be Classical T Tauri Stars \citep{Barentsen2011a}.
These are young objects
which are thought to show \ha\ in emission
due to the presence of hot, infalling gas
which is accreting onto the star
from a circumstellar disk.
This is likely to be the case for the candidate \ha-emitters
we discovered towards Sh 2-82 as well.
Two of our candidates, \#1 and \#4 in Table~\ref{tbl:emitters},
have previously been identified
as candidate YSOs by \cite{Robitaille2008}
and \cite{Yu2012}, respectively.
In these studies, the authors used Spitzer data
to find intrinsically red objects,
with SEDs consistent with the presence of a circumstellar disk.
Although the other four candidate emitters in our sample
have not previously appeared in the literature,
we note that all four are detected
in the Spitzer 8.0~\micron\ image at S/N$>$5.
They are likely to be YSOs exhibiting a mild infrared excess.

Sh 2-82 is one of a large population 
of poorly-studied star-forming regions
located at low Galactic latitudes,
which have only recently started to become revealed
by efforts to catalogue the wealth of `bubbles' detected
at mid-infrared wavelengths \citep{Churchwell2006,Simpson2012},
and by efforts to catalogue previously unknown clusters seen 
in the near-infrared \cite[e.g.][]{Bica2003}.
IPHAS data can offer a handle
on the extinction, distance
and stellar contents of many of such unexplored regions.
{\bf For example,
in the region studied above,
the IPHAS catalogue and the Spitzer/GLIMPSE archive
have 4798 sources in common
within a cross-matching radius of 1 arcsec,
(out of a total of 10\,739 and 11\,321 entries in the
IPHAS and Spitzer catalogues, respectively).
This is a favourable statistic,
given that this is a high-extinction sightline towards the inner plane
where an optical survey will naturally be extinction limited,
and IPHAS can be shown not to reach beyond 4-5\,kpc \citep{Sale2014}.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.

>>>>

\begin{figure}
    \includegraphics[width=\linewidth]{figures/fieldpair/fieldpair.pdf} 
    \caption{
    Example footprint of a pointing and its offset position.
    Each field in the survey is accompanied by an offset field
    at 5\arcmin\ West and South
	to deal with gaps between the CCDs.
	This unit of observation is called a \emph{field pair}.
	The WFC is a mosaic of four CCDs,
	and hence a field pair is composed of eight CCD frames.
	In this example, each CCD is plotted in semi-transparent grey
	to highlight the overlap regions.
	}
    \label{fig:fieldpair}
\end{figure}

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks.
An example of the footprint of a pointing
with its offset position is shown in Fig.~\ref{fig:fieldpair}.
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

Having carried out object detection,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
which uses a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the field distortion of the WFC.
The parameters of this solution
are then progressively refined by fitting against the
Two-Micron All Sky Survey \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector
\citep{Gonzalez-Solares2008}.

An external validation of the astrometry 
has been carried out by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the position offsets against UCAC4
computed for 1.3 million stars
in the magnitude range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean position offset to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.
We note that both UCAC4 and 2MASS -- and hence also IPHAS --
are calibrated in the International Celestial Reference System (ICRS).
%(Figure: note that blended stars were excluded).
%We note that the mean residual is  
%is $-0.006 \pm 0.086$ in R.A.
%and $0.012\pm0.073$ in Dec.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of the distance
    		between the positions of stars
    		which appear both in IPHAS DR2 and UCAC4
    		within a cross-matching distance of 1\arcsec.
            The residuals were computed for 
            the 1.3 million stars in the IPHAS catalogue
            which are not blended, not saturated,
            and fall in the magnitude range $13 < r < 15$.}
    \label{fig:astrometry}
\end{figure}

@@

In previous work, we have shown that the majority of
\ha-emitters seen by IPHAS towards an H{\sc ii} region
are likely to be Classical T Tauri Stars \citep{Barentsen2011a}.
These are young objects
which are thought to show \ha\ in emission
due to the presence of hot, infalling gas
which is accreting onto the star
from a circumstellar disk.
This is likely to be the case for the candidate \ha-emitters
we discovered towards Sh 2-82 as well.
Two of our candidates, \#1 and \#4 in Table~\ref{tbl:emitters},
have previously been identified
as candidate YSOs by \cite{Robitaille2008}
and \cite{Yu2012}, respectively.
In these studies, the authors used Spitzer data
to find intrinsically red objects,
with SEDs consistent with the presence of a circumstellar disk.
Although the other four candidate emitters in our sample
have not previously appeared in the literature,
we note that all four are detected
in the Spitzer 8.0~\micron\ image at S/N$>$5.
They are likely to be YSOs exhibiting a mild infrared excess.
The recovery of \ha-emitters in Spitzer
data illustrates how IPHAS can complement infrared surveys.

\subsection{IPHAS as a complement to infrared data}

{\bf Towards Sh2-82,
the IPHAS catalogue and the Spitzer/GLIMPSE archive
have 4798 entries in common,
out of a total of 10\,739 and 11\,321 entries, respectively.
1356 objects of these objects were detected in all
four GLIMPSE bands.
Fig.~\ref{fig:spitzer-ccd} presents their colour-colour diagram,
in which we have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles),
as well as the 511 objects that don't (red crosses).

\begin{figure}
    \includegraphics[width=\linewidth]{figures/spitzer-ccd/spitzer-ccd.pdf} 
    \caption{Spitzer/GLIMPSE colour-colour diagram towards Sh~2-82,
    showing the objects for which a counterpart exists in IPHAS
    (845 objects, grey circles) 
    and those for which no counterpart was identified 
    (511 objects, red crosses).
    A cross-matching radius of 1 arcsec
    was used to identify counterparts.
    The arrow illustrates the reddening vector corresponding
    to $A_{Ks}=3$ following the reddening law due to \citet{Flaherty2007}.
    It is apparent that the most heavily reddened objects
    tend to have no counterpart.
    Three heavily-reddened objects without counterpart
    are located above the diagram but are not shown for clarity. }
    \label{fig:spitzer-ccd}
\end{figure}

We find that objects located in the upper half of the diagram,
where highly-reddened objects are expected to sit,
are less likely to appear in both surveys.
This is not surprising,
because Sh2-82 is located 
towards a high-extinction sightline 
near the inner plane
where an optical survey will naturally be extinction-limited.
Nevertheless, the extinction map due to \citet{Sale2014}
has demonstrated that IPHAS can probe stars as far away 
as 4-5\,kpc towards this region.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.

Finally, we note that Sh 2-82 is one of a large population 
of poorly-studied star-forming regions
located at low Galactic latitudes,
which have only recently started to become revealed
by efforts to catalogue the wealth of `bubbles' detected
at mid-infrared wavelengths \citep{Churchwell2006,Simpson2012},
and by efforts to catalogue previously unknown clusters seen 
in the near-infrared \cite[e.g.][]{Bica2003}.
IPHAS data can offer a handle
on the extinction, distance
and stellar contents of many of such unexplored regions.
}
